Synthesis of Linear Functional Programs

### 2021-07-13 article/article.tex
notes

However, with the addition of recursion and
richer types to the synthesis and language, formal guidelines
\todo{nao sao formal guidelines, nao existem trabalhos a combinar
  estas features todas, essa parte é importante. Além de que 
ao incluir estas coisas deixamos de estar completamente em proof
search, pois muitas destas coisas (e.g. recursao) nao sao
necessariamente sound logicamente.}
  Tenta usar termos 
  especificos. Richer types? Richer como?} to support
the implementation become scarcer, or more complex \todo{nao existem
  ou sao complicados? ambos? citacao?}. Accordingly,
rather than derived from explicit rules, solutions appear first to
tame the extensions' inherent complex problems (such as infinite
recursion), and implicit rules come only as a byproduct. \todo{ Nao
  percebo o que queres dizer aqui. O que sao explicit rules e implicit
  rules? Ou há regras ou nao há. O que estas a dizer é que inventaste
  tecnicas de sintese para estes casos, que correspondem as regras que
  vais apresentar. E tao explicito ou implicito como o que esta na literatura.}

@@

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing. As explained in\ref{sec:background}, focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synth a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output\ref{sec:background} linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x \rightarrow M$ and the output linear context is
$\Delta'$.

@@

The decision rules are naturally followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either instance
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Ahead, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type instanceable by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: $\textrm{data Container} =
\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.

The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:
\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
\lolli C$, and its derivation:
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of Bar}\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of Bar}\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of Bar}\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    \te{Foo} \lolli C \Uparrow}
\end{mathpar}
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Simetrically, the
derivation for the synth goal $\te{Foo}$ doesn't have an end either.
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
    \Downarrow}
\end{mathpar}
To handle this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
for restrictions on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
These modifications prevent the derivation from not ending for the two seen
situations, however, they also greatly limit the derivable valid programs space,
leaving the synthetiser pretty much unable to synth specifications with
recursive types, which is so undesirable that we add three rules to complement the
restrictions on construction and destruction of recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra
The ADT rules draw closer to a final form, but will still need a revision
after the introduction of polymorfism and polymorfic parametrization on ADTs.

@@

\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
bound variables that quantify universally the type variables in $X$, called a
\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
First, \textbf{type variables are considered atomic types}, then, instantiate
the bound vars of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
$\alpha0 \lolli \alpha0$. We add a rule for this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
a type variable can instance a goal that's a type variable of the same name),
corresponds to the synthesis of a program where some expressions are treated
agnostically -- nothing constrains their type, meaning the program is
polymorphic * TODO: rewrite this?.

>>>>

However, with the addition of recursion and
richer types to the synthesis and language, formal guidelines
\todo{nao sao formal guidelines, nao existem trabalhos a combinar
  estas features todas, essa parte é importante. Além de que 
ao incluir estas coisas deixamos de estar completamente em proof
search, pois muitas destas coisas (e.g. recursao) nao sao
necessariamente sound logicamente.
  Tenta usar termos 
  especificos. Richer types? Richer como?} to support
the implementation become scarcer, or more complex \todo{nao existem
  ou sao complicados? ambos? citacao?}. Accordingly,
rather than derived from explicit rules, solutions appear first to
tame the extensions' inherent complex problems (such as infinite
recursion), and implicit rules come only as a byproduct. \todo{ Nao
  percebo o que queres dizer aqui. O que sao explicit rules e implicit
  rules? Ou há regras ou nao há. O que estas a dizer é que inventaste
  tecnicas de sintese para estes casos, que correspondem as regras que
  vais apresentar. E tao explicito ou implicito como o que esta na literatura.}

@@

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing. As
explained in\ref{sec:background} \todo{se o background nao vai ser
  incluido no documento principal, nao deves referenciar assim tao
  directamente para o conteudo principal.}, focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

\todo{Primeiro explica os varios judgments (as ``formas'' das
  conclusoes das regras) e o que significam as varias componentes,
  nomeadamente a parte da gestao de recursos.  Depois podes explicar
  regras concretas como fazes a seguir. Tambem seria util dizer quais
  os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
  faz ideia do que estas a dizer.}

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synth a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x \rightarrow M$ and the output linear context is
$\Delta'$.

\todo{a sintaxe $\lambda x \rightarrow M$ nao e mt standard, usa
  $\lambda x . M$}

@@

The decision rules are naturally followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either
instance \todo{instantiate}
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type that can be instantiated by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: $\textrm{data Container} =
\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.

The semantics of the ADTs relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:

\todo[inline]{Não uses Foo e Bar. $T_1$, $T_2$ , $\dots$ para
  tipos. $C_1$, $C_2$, $\dots$ para os construtores.}

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs \todo{ Na escrita, repetition
  does not legitimize. Constructors can take as arguments values of
  the type they are defining, ou algo assim}. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
\lolli C$, and its derivation:
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of Bar}\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of Bar}\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of Bar}\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    \te{Foo} \lolli C \Uparrow}
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $\te{Foo}$ is also infinite.
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
    \Downarrow}
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. \todo{Não se percebe muito bem a
  segunda parte da frase, rephrase!}
For this, we need two more contexts, $\Rho_C$
for restrictions \todo{constraints} on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
These modifications prevent the derivation from not ending
\todo{infinite derivations} for the scenarios described above,
however, they also greatly limit the space of derivable programs,
leaving the synthetiser effectively unable to synth specifications
with recursive types\todo{refrasear tendo em conta que se sintetiza a
  partir de specs e nao specs}, which is so undesirable that we add three rules
to complement the restrictions on construction and destruction of
recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

\todo[inline]{esta frase nao me parece muito necessaria}
The ADT rules draw closer to a final form, but will still need a revision
after the introduction of polymorfism and polymorfic parametrization on ADTs.

@@

\todo[inline]{Polymorphic}
